On Sample Size Needed for Block Bootstrap Confidence Intervals to Have Desired Coverage Rates

### 2024-03-29 Manuscript/manuscript.tex
ajur style
reference fix

% This designs the title display style for the maketitle command
\makeatletter
\newcommand\sixteen{\@setfontsize\sixteen{17pt}{6}}
\renewcommand{\maketitle}{\bgroup\setlength{\parindent}{0pt}
\begin{flushleft}
\sixteen\bfseries \@title
\medskip
\end{flushleft}
\textit{\@author}
\egroup}
\makeatother

@@

% Add author information below. Communicating author is indicated by an 
% asterisk, the affiliation is shown by superscripted lower case letter if 
% several affiliations need to be noted.
\author{Mathew Chandy*, Elizabeth D. Schifano, Jun Yan \\ \medskip 
Department of Statistics, University of Connecticut, Storrs, CT \\ \medskip 
Student: mathew.chandy@uconn.edu* \\
Mentors: elizabeth.schifano@uconn.edu, jun.yan@uconn.edu
}
\pagestyle{empty}
\begin{document}

@@

\section*{Block Bootstrap CIs}
\label{sec:bbci}

@@

We generated time series $X_t$ from a 1st 
order autoregressive (AR(1)) process:
\begin{equation*}
X_t = \phi X_{t-1} + \epsilon_t,
\end{equation*}
where $\phi$ is an autoregressive coefficient, and $\epsilon_t$ is a series of
independent errors from a normal distribution with mean zero and variance
$\sigma_{\epsilon}^2$. The strength of the serial dependence is controlled by
$\phi$, which was set to five levels: $\{-0.4, -0.2, \allowbreak 0.0, 0.2, 0.4\}$. 
We only
used serial dependences as strong as 0.4, because we only seek to 
establish the 
general trend as the strength of the autocorrelation 
increases, and how it varies depending on the sign of the autocorrelation and 
the parameter of interest. The 
series $X_t$ has mean zero and variance 
$\sigma_x^2 = (1 - \phi^2) \sigma_{\epsilon}^2$, so for each value of $\phi$, we 
set $\sigma_{\epsilon}^2 = 1 / (1 - \phi^2)$ such that $\sigma_x^2 = 1$.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=\textwidth]{figures/alt_plot_norm1}
  \caption{Empirical coverage rates of different 95\% block bootstrap CIs for
    the marginal mean $\mu$, the marginal standard deviation $\sigma_x$, and
    the first-order autocorrelation coefficient $\phi$ of an AR(1) process with 
    a marginal standard 
    normal distribution with AR coefficient
    $\phi \in \{-0.4, 0.2, 0, 0.2, 0.4\}$ and series length
    $n \in \{100, 200, 400, 800, 1600, 3200\}$ based on 10,000 replicates
    of
    block bootstrap with $l = \lceil n^{1/3} \rceil$. The
    error bars represent 95\% CIs of the real coverage rates.}
  \label{fig:norm1}
\end{figure}

For estimating the mean parameter~$\mu$, \textbf{Figure~\ref{fig:norm1}} suggests 
that all methods eventually approach correct coverage of $\mu$ as sample size 
increases. Student's $t$ CIs appear to need the smallest sample size to achieve 
correct coverage, except for samples with strong negative dependence, in which 
case, they actually over-cover $\mu$ for smaller sample sizes. For instance, for 
a sample with $n = 100$ and $\phi = -0.4$, the lower bound for a Students $t$ 
CI's coverage of $\mu$ is greater than 95\%, whereas the coverage intervals for 
other methods contain 95\%. The standard normal, percentile, BC, and BCA, and 
recentered percentile CIs require similar sample sizes to recover $\mu$ at the 
nominal level for all combinations of $n$ and $\phi$. All methods seem to 
require a smaller sample to recover $\mu$ at the nominal rate when dealing with 
negative dependence versus positive dependence. For example, BC CIs recover 
$\mu$ for $n \geq 100$ when $\phi = -0.2$, but they only recover $\mu$ for 
$n \geq 800$ when $\phi = 0.2$. In addition, as a negative dependence gets 
stronger, holding everything else equal, coverage increases, which lead to the 
Student $t$ CI's aforementioned over-coverage. As a positive dependence gets 
stronger, holding everything else equal, coverage decreases, and a larger sample 
is necessary to recover~$\mu$. A possible explanation for this is that if a 
stationary series has a positive 
autocorrelation, the effective sample
size is decreased, whereas if a series has a negative autocorrelation, the
effective sample size is increased.\citep{geyer2011introduction}
Additionally,  this seems to have a 
greater effect on the the estimation of the location parameter versus that
of the scale parameter or temporal dependence parameter.

@@

The results for $l = \lceil 2n^{1/3} \rceil$ are reported in the Appendix.
Performance generally seems to be better for $l = \lceil n^{1/3} \rceil$, but
importantly, the patterns in performance when varying other parameters
appear to be robust to the different block size.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=\textwidth]{figures/alt_plot_exp1}
  \caption{Empirical coverage rates of different 95\% block bootstrap CIs for
    the marginal mean $\mu$, the marginal standard deviation $\sigma_w$, and
    the first-order autocorrelation coefficient $\rho$ of a stationary series 
    with 
    marginal unit exponential distribution obtained by transforming an AR(1) 
    process
    with $\phi \in \{-0.4, 0.2, 0, 0.2, 0.4\}$ with series length 
    $n \in \{100, 200, 400, 800, 1600, 3200\}$ based on 10,000 replicates 
    replicates of
    block bootstrap with $l = \lceil n^{1/3} \rceil$. 
    The error bars represent 95\% CIs of the real coverage rates.}
  \label{fig:exp1}
\end{figure}

For the scenario of marginal exponential distribution, the empirical
coverage rates for $\mu$, $\sigma_w$, and 
the lag-1 autocorrelation coefficient $\rho$ using block bootstrap
with $l = \lceil n^{1/3} \rceil$, as well as 95\%
confidence intervals of the real coverage are displayed in 
\textbf{Figure~\ref{fig:exp1}}. 
Additionally, a set of 50 CIs are displayed for each of the percentile, 
BC, and recentered percentile approaches for exponentially distributed
samples of $n \in \{800, 1600\}$ with lag-1 autocorrelation coefficient 0.355
($\phi = 0.4$) in \textbf{Figure~\ref{fig:eri}}.

@@

It appears that a greater sample size is generally required for the bootstrap
CIs to cover the mean and standard deviation parameters in the exponential
margin case than in the normal margin case. However, the other trends 
and patterns discussed regarding the performance of various methods and
diverse parameters remain unchanged. For example, Student's $t$ confidence
intervals still exhibit higher coverage rates in comparison to alternative
methods. Performance continues to be more favorable when temporal dependence is
negative rather than positive. Of particular significance, the percentile, BC,
and BCA confidence intervals still display a decline in coverage accuracy for
the lag-1 autocorrelation coefficient as sample size increases as demonstrated
in Figure~\ref{fig:eri}. Both the percentile and BC intervals persist in
manifesting the same bias issue. On the other hand, the recentered percentile
confidence interval continues to be effective in estimating the temporal 
dependence due to the
inherent unbiasedness of the original point estimator. In sum, 
for the most part, the findings for
series that are marginally exponentially distributed closely mirror those
attained for series that are marginally normally distributed.

@@

Block bootstrap is a useful method for estimating parameters of a time series,
from simple parameters like the mean to more complicated temporal dependence
factors. We know theoretically that the block bootstrap procedure will cover a 
parameter of a time series at the nominal level given an infinitely large
%\jy{Also cite Calhoun in the introduction}\mc{addressed}
sample,\citep{calhoun2018} so the goal for this study was to find the smallest 
finite sample length~$n$
of a time series in order for the block bootstrap procedure to recover its 
associated parameters at an acceptable rate. Our analysis relies on the 
assumption that there is a size $n$ large enough for the method to work: that 
is, the method's performance improves as $n$ increases. Out of the six types of 
intervals used in this study, this assumption was found to hold true with 
respect to estimating $\phi$ only for standard normal, Student's $t$, and recentered 
percentile CIs, whereas percentile, BC, and BCA intervals exhibited coverage 
deterioration as $n$ increased. The percentile CI's coverage deterioration can 
be attributed to bias that is not corrected as $n$ increases. Specifically, as 
$n$ increases, the width of the CI decreases, but because the percentile CI 
underestimates $\phi$, the coverage decreases. The BC CI seems to correct the 
bias, but the width of the CI seems to be too short. The acceleration factor 
of the BCA CI seems to fail, as the width of the CI seems to be too
short.

@@

\section*{Appendix: Simulation Study with $l = \lceil 2n^{1/3} \rceil$}

\textbf{Figure~\ref{fig:norm2}} 
summarizes the empirical coverage rates and the 95\% confidence intervals of the 
real coverage for a marginal standard normal distribution using block bootstrap
with $l = \lceil 2n^{1/3} \rceil$.

\begin{figure}[tbp]
  \centering
  \includegraphics[width=\textwidth]{figures/alt_plot_norm2}
  \caption{Empirical coverage rates of different 95\% block bootstrap CIs for
    the marginal mean $\mu$, the marginal standard deviation $\sigma_x$, and
    the first-order autocorrelation coefficient $\phi$ of an AR(1) process with 
    a marginal standard 
    normal distribution with AR coefficient
    $\phi \in \{-0.4, 0.2, 0, 0.2, 0.4\}$ and series length
    $n \in \{100, 200, 400, 800, 1600, 3200\}$ based on 10,000 replicates of
    block bootstrap with $l = \lceil 2n^{1/3} \rceil$. The
    error bars represent 95\% CIs of the real coverage rates.}
  \label{fig:norm2}
\end{figure}

Based on \textbf{Figure~\ref{fig:norm2}}, for negative autocorrelations, the 
coverage 
rates of $\mu$ appear to be lower when
using $l = \lceil 2n^{1/3} \rceil$. For example, whereas $n = 100$ or 200 
would seem sufficient for most CIs when using $l = \lceil n^{1/3} \rceil$, 
$n = 800$ or 1600 is necessary to capture negative autocorrelations for 
$l = \lceil n^{1/3} \rceil$. Student's $t$ CIs do not seem to be as affected
by this change in $l$: for $\phi = -0.4$ and -0.2, they still over-cover $\mu$
for smaller values of~$n$. The results for $\sigma_x$ with $l = \lceil 2n^{1/3} 
\rceil$ look very similar to 
those the results for $\sigma_x$ with $l = \lceil n^{1/3} \rceil$, but coverage 
rates of $\sigma_x$ do look slightly lower
especially for negative values of $\phi$, although Student's $t$ CIs are again
not as influenced by this change in~$l$. A larger sample size seems necessary
when using other CIs to estimate $\sigma_x$ for $l = \lceil 2n^{1/3} \rceil$.
%\jy{Put the small paragraphs into a bigger one.}\mc{addressed}
Recentered percentile and standard CIs have slightly lower coverage rates when
estimating negative values of $\phi$ with $l = \lceil 2n^{1/3} \rceil$. 
Although it is still a problem, the coverage deterioration appears to be less 
dramatic for BCA,
BC, and percentile CIs. Aside from these differences, the overall changes
in performance when other experimental factors are changed are the same as
when $l = \lceil n^{1/3} \rceil$.

\begin{figure}[tbp]
  \centering
  \includegraphics[width=\textwidth]{figures/alt_plot_exp2}
  \caption{Empirical coverage rates of different 95\% block bootstrap CIs for
    the marginal mean $\mu$, the marginal standard deviation $\sigma_w$, and
    the first-order autocorrelation coefficient $\rho$ of a stationary series 
    with 
    marginal unit exponential distribution obtained by transforming an AR(1) 
    process
    with $\phi \in \{-0.4, 0.2, 0, 0.2, 0.4\}$ with series length 
    $n \in \{100, 200, 400, 800, 1600, 3200\}$ based on 10,000 replicates 
    replicates of
    block bootstrap with $l = \lceil 2n^{1/3} \rceil$. 
    The error bars represent 95\% CIs of the real coverage rates.}
  \label{fig:exp2}
\end{figure}

\textbf{Figure~\ref{fig:exp2}} 
summarizes the empirical coverage rates and the 95\% confidence intervals of the 
real coverage for a marginal exponential distribution using block bootstrap
with $l = \lceil 2n^{1/3} \rceil$. Again, altering the block size results in the same
changes in performance of different CIs as those in the scenario of
marginal normal distribution.

\bibliographystyle{unsrtnat}
\bibliography{citations}

\section*{About Student Author}
\label{sec:asa}
Mathew Chandy is a senior majoring in both Statistics and Statistical Data Science, 
and he plans to graduate in the spring of 2024.

>>>>

% This designs the title display style for the maketitle command
\makeatletter
\newcommand\sixteen{\@setfontsize\sixteen{16pt}{6}}
\renewcommand{\maketitle}{\bgroup\setlength{\parindent}{0pt}
\begin{flushleft}
\sixteen\bfseries \@title
\medskip
\end{flushleft}
\textit{\@author}
\egroup}
\makeatother

@@

% Add author information below. Communicating author is indicated by an 
% asterisk, the affiliation is shown by superscripted lower case letter if 
% several affiliations need to be noted.
\author{Mathew Chandy*, Elizabeth D. Schifano, \& Jun Yan \\ \medskip \medskip 
Department of Statistics, University of Connecticut, Storrs, CT \\ \medskip\medskip 
https://doi.org/10.33697/ajur.2024.101\\ \medskip\medskip 
Student: mathew.chandy@uconn.edu* \\
Mentors: elizabeth.schifano@uconn.edu, jun.yan@uconn.edu
}
\pagestyle{empty}
\begin{document}

@@

\section*{\textit{Block Bootstrap CIs}}
\label{sec:bbci}

@@

We generated time series $X_t$ from a 1st 
order autoregressive (AR(1)) process:
\begin{equation*}
X_t = \phi X_{t-1} + \epsilon_t,
\end{equation*}
where $\phi$ is an autoregressive coefficient, and $\epsilon_t$ is a series of
independent errors from a normal distribution with mean zero and variance
$\sigma_{\epsilon}^2$. The strength of the serial dependence is controlled by
$\phi$, which was set to five levels: $\{-0.4, -0.2, \allowbreak 0.0, 0.2, 0.4\}$. 
We only
used serial dependences as strong as 0.4, because we only seek to 
establish the 
general trend as the strength of the autocorrelation 
increases, and how it varies depending on the sign of the autocorrelation
and 
the parameter of interest. The 
series $X_t$ has mean zero and variance 
$\sigma_x^2 = \sigma_{\epsilon}^2 / (1 - \phi^2)$, so for each value of
$\phi$, 
we 
set $\sigma_{\epsilon}^2 = (1 - \phi^2)$ such that $\sigma_x^2 = 1$.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=\textwidth]{figures/alt_plot_norm1}
  \includegraphics[width=\textwidth]{figures/alt_plot_norm2}
  \caption{Empirical coverage rates of different 95\% block bootstrap CIs for
    the marginal mean $\mu$, the marginal standard deviation $\sigma_x$, and
    the first-order autocorrelation coefficient $\phi$ of an AR(1) process with 
    a marginal standard 
    normal distribution with AR coefficient
    $\phi \in \{-0.4, 0.2, 0, 0.2, 0.4\}$ and series length
    $n \in \{100, 200, 400, 800, 1600, 3200\}$ based on 10,000 replicates
    of
    block bootstrap with $l = \lceil n^{1/3} \rceil$. The
    error bars represent 95\% CIs of the real coverage rates.
  Top: $l = \lceil n^{1/3} \rceil$. Bottom: $l = \lceil 2n^{1/3} \rceil$.}
  \label{fig:norm1}
\end{figure}

For estimating the mean parameter~$\mu$, the top panel of
\textbf{Figure~\ref{fig:norm1}} suggests 
that all methods eventually approach correct coverage of $\mu$ as sample size 
increases. Student's $t$ CIs appear to need the smallest sample size to achieve 
correct coverage, except for samples with strong negative dependence, in which 
case, they actually over-cover $\mu$ for smaller sample sizes. For instance, for 
a sample with $n = 100$ and $\phi = -0.4$, the lower bound for a Students $t$ 
CI's coverage of $\mu$ is greater than 95\%, whereas the coverage intervals for 
other methods contain 95\%. The standard normal, percentile, BC, and BCA, and 
recentered percentile CIs require similar sample sizes to recover $\mu$ at the 
nominal level for all combinations of $n$ and $\phi$. All methods seem to 
require a smaller sample to recover $\mu$ at the nominal rate when dealing with 
negative dependence versus positive dependence. For example, BC CIs recover 
$\mu$ for $n \geq 100$ when $\phi = -0.2$, but they only recover $\mu$ for 
$n \geq 800$ when $\phi = 0.2$. In addition, as a negative dependence gets 
stronger, holding everything else equal, coverage increases, which lead to the 
Student $t$ CI's aforementioned over-coverage. As a positive dependence gets 
stronger, holding everything else equal, coverage decreases, and a larger sample 
is necessary to recover~$\mu$. A possible explanation for this is that if a 
stationary series has a positive 
autocorrelation, the effective sample
size is decreased, whereas if a series has a negative autocorrelation, the
effective sample size is increased.\citep{geyer2011introduction}
Additionally,  this seems to have a 
greater effect on the the estimation of the location parameter versus that
of the scale parameter or temporal dependence parameter.

@@

The results for $l = \lceil 2n^{1/3} \rceil$ are reported in the
bottom panel of Figure~\ref{fig:norm1}. The performances generally
seems to be inferior compared those with $l = \lceil n^{1/3}\rceil$, but
importantly, the patterns in performance when varying other parameters
appear to be robust to the different block size.
For negative autocorrelations, the coverage 
rates of $\mu$ appear to be lower when
using $l = \lceil 2n^{1/3} \rceil$. For example, whereas $n = 100$ or 200 
would seem sufficient for most CIs when using $l = \lceil n^{1/3} \rceil$, 
$n = 800$ or 1600 is necessary to capture negative autocorrelations for 
$l = \lceil n^{1/3} \rceil$. Student's $t$ CIs do not seem to be as affected
by this change in $l$: for $\phi = -0.4$ and -0.2, they still over-cover $\mu$
for smaller values of~$n$. The results for $\sigma_x$ with $l = \lceil 2n^{1/3} 
\rceil$ look very similar to 
those the results for $\sigma_x$ with $l = \lceil n^{1/3} \rceil$, but coverage 
rates of $\sigma_x$ do look slightly lower
especially for negative values of $\phi$, although Student's $t$ CIs are again
not as influenced by this change in~$l$. A larger sample size seems necessary
when using other CIs to estimate $\sigma_x$ for $l = \lceil 2n^{1/3} \rceil$.
Recentered percentile and standard CIs have slightly lower coverage rates when
estimating negative values of $\phi$ with $l = \lceil 2n^{1/3} \rceil$. 
Although it is still a problem, the coverage deterioration appears to be less 
dramatic for BCA,
BC, and percentile CIs. Aside from these differences, the overall changes
in performance when other experimental factors are changed are the same as
when $l = \lceil n^{1/3} \rceil$.

@@

\begin{figure}[tbp]
  \centering
  \includegraphics[width=\textwidth]{figures/alt_plot_exp1}
  \includegraphics[width=\textwidth]{figures/alt_plot_exp2}
  \caption{Empirical coverage rates of different 95\% block bootstrap CIs for
    the marginal mean $\mu$, the marginal standard deviation $\sigma_w$, and
    the first-order autocorrelation coefficient $\rho$ of a stationary series 
    with 
    marginal unit exponential distribution obtained by transforming an AR(1) 
    process
    with $\phi \in \{-0.4, 0.2, 0, 0.2, 0.4\}$ with series length 
    $n \in \{100, 200, 400, 800, 1600, 3200\}$ based on 10,000 replicates 
    replicates of
    block bootstrap with $l = \lceil n^{1/3} \rceil$. 
    The error bars represent 95\% CIs of the real coverage rates.
    Top: $l = \lceil n^{1/3} \rceil$. Bottom: $l = \lceil 2n^{1/3} \rceil$.}
  \label{fig:exp1}
\end{figure}

For the scenario of marginal exponential distribution, the empirical
coverage rates for $\mu$, $\sigma_w$, and 
the lag-1 autocorrelation coefficient $\rho$ using block bootstrap
with $l  \in \{\lceil n^{1/3} \rceil, \lceil 2n^{1/3} \rceil$, as well as 95\%
confidence intervals of the real coverage are displayed in 
\textbf{Figure~\ref{fig:exp1}}. 
Additionally, a set of 50 CIs are displayed for each of the percentile, 
BC, and recentered percentile approaches for exponentially distributed
samples of $n \in \{800, 1600\}$ with lag-1 autocorrelation coefficient 0.355
($\phi = 0.4$) in \textbf{Figure~\ref{fig:eri}}.

@@

It appears that a greater sample size is generally required for the bootstrap
CIs to cover the mean and standard deviation parameters in the exponential
margin case than in the normal margin case. However, the other trends 
and patterns discussed regarding the performance of various methods and
diverse parameters remain unchanged. For example, Student's $t$ confidence
intervals still exhibit higher coverage rates in comparison to alternative
methods. Performance continues to be more favorable when temporal dependence is
negative rather than positive. Again, altering the block size results
in the same changes in performance of different CIs as those in the
scenario of  marginal normal distribution.
Of particular significance, the percentile, BC,
and BCA confidence intervals still display a decline in coverage accuracy for
the lag-1 autocorrelation coefficient as sample size increases as demonstrated
in Figure~\ref{fig:eri}. Both the percentile and BC intervals persist in
manifesting the same bias issue. On the other hand, the recentered percentile
confidence interval continues to be effective in estimating the temporal 
dependence due to the
inherent unbiasedness of the original point estimator. In sum, 
for the most part, the findings for
series that are marginally exponentially distributed closely mirror those
attained for series that are marginally normally distributed.

@@

Block bootstrap is a useful method for estimating parameters of a time series,
from simple parameters like the mean to more complicated temporal dependence
factors. We know theoretically that the block bootstrap procedure will cover a 
parameter of a time series at the nominal level given an infinitely large
sample,\citep{calhoun2018} so the goal for this study was to find the smallest 
finite sample length~$n$
of a time series in order for the block bootstrap procedure to recover its 
associated parameters at an acceptable rate. Our analysis relies on the 
assumption that there is a size $n$ large enough for the method to work: that 
is, the method's performance improves as $n$ increases. Out of the six types of 
intervals used in this study, this assumption was found to hold true with 
respect to estimating $\phi$ only for standard normal, Student's $t$, and recentered 
percentile CIs, whereas percentile, BC, and BCA intervals exhibited coverage 
deterioration as $n$ increased. The percentile CI's coverage deterioration can 
be attributed to bias that is not corrected as $n$ increases. Specifically, as 
$n$ increases, the width of the CI decreases, but because the percentile CI 
underestimates $\phi$, the coverage decreases. The BC CI seems to correct the 
bias, but the width of the CI seems to be too short. The acceleration factor 
of the BCA CI seems to fail, as the width of the CI seems to be too
short.

@@

\bibliographystyle{ajur}
\input{citation}

\section*{About Student Author}
\label{sec:asa}
Mathew Chandy is a senior majoring in both Statistics and Statistical Data Science, 
and he plans to graduate in the Spring of 2024.
